\begin{longtable}{p{3cm}p{10cm}}  % Begins a longtable environment. The 'p{3cm}' and 'p{10cm}' specify column widths.\\
AAII & Artificial Analysis Intelligence Index. An index used to benchmark and compare the performance and capabilities of different artificial intelligence models.\\
API & Application Programming Interface. A software intermediary that allows different applications or services to communicate and exchange data with one another.\\
AST & Abstract Syntax Tree. A tree representation of the structural logic of source code, commonly used in static analysis to examine how a program is built.\\
BLOB & Binary Large Object. A database data type used to store massive amounts of unstructured data, such as images, audio files, or videos.\\
CI & Continuous Integration. A software development practice where developers frequently and automatically merge their code changes into a central repository.\\
CLI & Command-Line Interface. A text-based interface used to interact with software by typing commands into a terminal.\\
Code Smell & A symptom in the source code of a computer program that possibly indicates a deeper problem or weakness in design, often used interchangeably with antipatterns.\\
Cohen’s Kappa & A statistical measure used to calculate inter-rater (or intra-rater) agreement for categorical items, accounting for agreement occurring by chance.\\
Context Rot & A phenomenon where a Large Language Model's performance consistently degrades and becomes increasingly unreliable as the input context length grows.\\
CoT & Chain-of-Thought. A prompting strategy for Large Language Models that guides the AI to articulate its reasoning step-by-step before providing a final answer.\\
CSV & Comma-Separated Values. A simple text file format widely used to store and transfer tabular data.\\
CV & Coefficient of Variation. A statistical measure that represents the ratio of the standard deviation to the mean, used to show the extent of variability in relation to the average of the population.\\
DAO & Data Access Object. A software design pattern that isolates the database interaction logic from the rest of the application's business logic.\\
DBMS & Database Management System. A software application used to create, manage, and securely handle data within databases.\\
DDL & Data Definition Language. A subset of SQL used to create, modify, and define database objects like tables.\\
DML & Data Manipulation Language. A subset of SQL used for adding, updating, and deleting data within a database.\\
DQL & Data Query Language. A subset of SQL used for searching and retrieving specific data from a database.\\
DSL & Domain-Specific Language. A specialised programming language created for a specific problem or context (e.g., jOOQ's built-in query language).\\
Enum & Enumeration. A data type consisting of a set of named values, used in programming and database schemas to restrict a variable or column to predefined choices.\\
F1-Score & A machine learning evaluation metric that measures a model's performance by balancing both precision and recall into a single harmonic mean score.\\
Few-Shot \newline Prompting & A technique where a Large Language Model is provided with a few examples of the desired task in the prompt to improve its output accuracy.\\
FN & False Negative. A detection error where an analysis tool fails to identify a flaw that actually exists in the code.\\
FP & False Positive. A detection error where an analysis tool mistakenly identifies a correct piece of code as flawed.\\
GUI & Graphical User Interface. A user interface that allows interaction with a program through visual elements like icons, buttons, and windows rather than text commands.\\
I/O & Input/Output. In the context of processing with language models, it refers to the data sent to the model (input tokens) and the response generated by it (output tokens).\\
IDE & Integrated Development Environment. A software application that provides developers with a comprehensive set of tools for writing, editing, debugging, and testing code.\\
IoU & Intersection over Union. An evaluation metric that measures the overlap between a predicted location and the actual target location, used in this thesis to assess antipattern localisation accuracy.\\
Jaccard Index & A statistical measure used to evaluate the similarity and diversity of sample sets. Applied in this thesis to measure the co-occurrence of different antipatterns.\\
JDBC & Java Database Connectivity. A standard Java API that enables Java applications to connect to and interact with relational databases.\\
jOOQ & jOOQ Object Oriented Querying. A Java library that allows developers to build and execute SQL queries in a type-safe, object-oriented manner.\\
JSON & JavaScript Object Notation. A widely used, lightweight data-interchange format that is easy for humans to read and for machines to parse.\\
JVM & Java Virtual Machine. System software that provides a runtime environment enabling computers to run compiled Java programs regardless of the underlying hardware.\\
LLM & Large Language Model. An artificial intelligence model trained on vast amounts of data, capable of understanding, analysing, and generating natural human language as well as programming code.\\
MENTOR & Measure, Explain, Nominate, Test, Optimize, Rebuild. A systematic framework and methodology used for evaluating and optimising the performance of database indexes.\\
ML & Machine Learning. A branch of artificial intelligence that enables systems to learn from past data and improve their performance on a specific task without being explicitly programmed.\\
Monte Carlo \newline Method & A broad class of computational algorithms that rely on repeated random sampling to obtain numerical results, used in this thesis for balancing dataset splits.\\
NMS & Non-Maximum Suppression. An algorithm used to filter out overlapping predictions by selecting the most probable bounding box or code span and discarding the rest.\\
ORM & Object-Relational Mapping. A programming technique that links relational database tables to application objects, hiding direct SQL code generation from the developer.\\
PL/SQL & Procedural Language/Structured Query Language. Oracle Corporation's procedural extension for SQL and the relational database.\\
POJO & Plain Old Java Object. An ordinary Java object that is not bound by any special restrictions or framework-specific rules.\\
Prompt \newline Engineering & The practice of designing and refining input instructions (prompts) to effectively guide Large Language Models toward generating specific, high-quality outputs.\\
Regex & Regular Expression. A sequence of characters that specifies a search pattern, often used for advanced text searching and manipulation.\\
SDK & Software Development Kit. A collection of software tools and libraries provided by hardware or software developers to build applications for specific platforms.\\
SotA & State-of-the-Art. The most recent, advanced, and highest-performing iteration of a technology or artificial intelligence model currently available.\\
Spearman's \newline Correlation & A statistical measure used to evaluate the strength and direction of the relationship between the rankings of two variables, showing if they tend to increase together.\\
SQL & Structured Query Language. A standardised programming language used for managing relational databases and performing various operations on the data within them.\\
SQL Antipattern & A frequently occurring, seemingly effective but ultimately counterproductive approach to database design or querying that degrades system performance, maintainability, or data integrity in the long run.\\
TN & True Negative. A correct identification where the analysis tool accurately determines that a specific code segment contains no flaws.\\
ToT & Tree-of-Thought. A prompting strategy for Large Language Models where the model explores and evaluates multiple reasoning branches simultaneously, simulating a discussion among subject-matter experts.\\
TP & True Positive. A correct identification where the analysis tool successfully finds and marks an actually existing flaw in the code.\\
VAT & Value Added Tax. A consumption tax placed on a product or service, relevant in this thesis for accurate API cost calculations.\\
Zero-Shot \newline Prompting & A technique where a Large Language Model is given a direct query or task instruction without any prior examples of correct answers, relying entirely on its pre-trained knowledge.\\
\end{longtable}
\addtocounter{table}{-1} % Decreases the table counter by 1 so that it doesn't increment the table number in the document